Impact of grids and dynamical cores in CESM2.2 on the surface mass balance of the Greenland Ice Sheet

### 2022-03-16 arcticgrids.tex
updates

General Circulation Models (GCMs) are powerful tools for understanding the meteorology and climate of the high-latitudes, which are among the most sensitive regions on Earth to global and environmental change. Despite their importance, the numerical treatment of polar regions in GCMs is handled in vastly-different ways due to the so-called \textit{pole-problem} \cite{W2007JMSJ}, which refers to numerical instability arising from the convergence of meridians at the polar points on latitude-longitude grids (e.g., Figure~\ref{fig:uni-grids}a). Depending on the numerics, methods exist to suppress this instability, and latitude-longitude grids may be advantageous for polar processes as structures can be represented with more degrees of freedom than elsewhere in the computational domain. With the recent trend towards globally uniform unstructured grids, any potential benefits of latitude-longitude grids on polar regions may become a relic of the past. In this study, a spectrum of model grids and dynamical cores (hereafter referred to as \textit{dycores}) available in the Community Earth System Model, version 2.2 (CESM; \url{http://www.cesm.ucar.edu/models/cesm2/}) are evaluated to understand their impacts on the simulated characteristics of the Arctic, with a special focus on the meteorology over the Greenland Ice Sheet.

@@

@@

All grids and dycores are run using an identical transient 1979-1998 AMIP-style configuration, with prescribed monthly SST/sea-ice after \cite{CESMSST}. This configuration refers to the $FHIST \ compset$ and runs out of the box in CESM2.2. The Community Land Model (CLM5; \url{https://escomp.github.io/ctsm-docs/versions/release-clm5.0/html/users_guide/index.html}), which uses the same grid as the atmosphere grid, calculates the surface energy balance at each land tile within a grid cell, which is used to compute snow and bare ice melting to inform the surface mass balance (SMB) of a glacier unit \cite{VKETAL2020JGR}. For land ice tiles coinciding with the Antarctic and Greenland Ice Sheets, the surface energy and mass balance are evaluated at ten elevation classes, and integrated over sub-grid hypsometric area-elevation bins for more accurate estimation of the surface mass balance of the big ice sheets. Ice accumulation is modeled as a capping flux, or snow in excess of the assumed 10 m snow cap, and refreezing of liquid within the snowpack additionally acts as a source of mass in the SMB calculation. Since the 10 m snowcap needs to be reached in the accumulation zone to simulate the SMB, the snow depths in the variable-resolution grids were spun-up  by forcing CLM5 standalone, cycling over 20 years of a fully coupled $ARCTIC$ run for about 500 years. The uniform resolution grids are all initialized with an SMB from an existing $f09$ spun-up initial condition.

@@

The SMB is the sum of the mass source term, accumulation (i.e., precipitation), and the mass sink term, ablation. Ablation can be expressed as sublimation plus total runoff, with runoff being a combination of liquid precipitation and snow and ice melt. Not all liquid precipitation becomes runoff; rain may penetrate pore spaces in the firn layer and freeze, forming ice lenses in the subsurface. All these processes are represented by different components in CESM, but the surface mass balance proper is computed by CLM. CLM uses a downscaling technique to account for sub-grid variability in SMB. In short, the ice sheet patch in a CLM grid cell is subdivided into 10 elevation classes (EC), weighted by their respective area fractions at each EC, which is derived from a high resolution GrIS elevation model. The near surface air temperature, humidity and air density are calculated at each EC using an assumed lapse rate and the elevation difference from the grid mean, allowing for the precipitation rates to be repartitioned into solid or liquid based on the temperature of each EC. This technique allows for a unique surface energy balance and surface mass balance at each EC. Integrating over all ECs using the area weights then provides a more accurate SMB. For a more detailed description of how the surface mass balance is computed in the CESM, see \cite{KETAL2020JAMES}.

@@

Simulated SMB are compared against several observational data products. RACMO2.3 11km and RACMO2.3p2 5.5km regional simulations forced by reanalysis at it's domain boundaries, performs very skillfully relative to observations and is therefore considered an ideal modeling target \cite{NETAL2019SCIENCE}. The Land Ice Verification and Validation toolkit (LIVVkit), version 2.1 \cite{LIVVkit} maintains a repository of snow pit and and ice core SMB measurements, as well as the IceBridge radar accumulation dataset.

A common high resolution dataset is used to generate the GrIS boundary conditions using the CLM dataset creation tools. Since we are interested in the total ice sheet SMB, we seek to integrate various components of the SMB over a common ice mask to get the total mass change over the GrIS. Figure~\ref{fig:grisdx} shows the GrIS ice mask area across the different grids, as a function of the number of grid points. Due to the use conservative regridding in the CLM tools, the interpolation errors are small and ice mask areas have less than $1.5\%$ errors relative to the raw ice mask dataset. RACMO, however, uses a smaller ice mask, about $3\%$ smaller than the raw CLM dataset. While Figure~\ref{fig:grisdx} suggests integrating quantities over the native ice mask of the six grids would probably not suffer from large errors due to differing ice masks, we seek to compare these integrated quantities with RACMO. Therefore, we have taken the approach of mapping all model fields to the lowest resolution grids and integrating over the respective low resolution ice masks. Due to the sensitivity of mapping errors to grid coordinates (i.e., unstructured or structured), all quantities are evaluated on both the $f19$ and $ne30pg2$ low resolution grids. In addition, two remapping packages are used; ESMF conservative and a TempestRemap high-order, monotone algorithm. In all, each integrated quantity is evaluated (at most) four times to provide an estimate of uncertainty due to differences in grid coordinates and remapping algorithm.

@@

Before delving into the simulated characteristics of the Arctic, the global mean differences between the various grids and dycores are assessed. Figure~\ref{fig:dT-lores} shows the climatological, zonal mean height plots expressed as differences between the uniform resolution grids and dycores. The $f09$ grid is warmer than the $f19$ grid, primarily in the mid-to-high latitudes and throughout the depth of the troposphere. This is a common response to increasing horizontal resolution in GCMs \cite{PS2002CD,RETAL2006JC}, and \cite{HK2020QJRMS} has shown that this occurs in CAM due to greater resolved vertical velocities that in turn, facilitate greater condensational heating in the macrophyiscs routine in CLUBB. The right columns in Figure~\ref{fig:dT-lores} supports this interpretation, which shows an increase in the climatological CLUBB heating in the mid-latitudes in the $f09$ grid.

@@

@@

Figure~\ref{fig:pointdiff} shows the distribution of point-wise differences from LIVVkit observational database, as a violin plot. The IceBridge dataset is exclusively from the interior of the ice sheet and represents accumulation rates. The uniform $1^{\circ}-2^{\circ}$ grids have similar median errors of about +35-50 mm.w.e, while the variable-resolution errors are noticeably smaller. The in-situ observations in the accumulation zone, shown in the middle plot looks very similar to the IceBridge errors, and so provides confidence that the variable-resolution grids are outperforming the uniform grids in the interior accumulation zone. The errors evaluated at in-situ ablation zone measurements are in tension with the RACMO results in Figure~\ref{fig:tseries}, indicating that the uniform $1^{\circ}-2^{\circ}$ grids perform similarly, the $ARCTIC$ grid showing a reduction in bias and the $ARCTICGRIS$ having the smallest bias of all the grids. However, the in-situ ablation measurements are sparse in time and space, and so there is a large uncertainty in extending these results to the entirety of the ablation zone.

The infamous k-transect mass balance profile is contained in the LIVVkit database, and shown in Figure~\ref{fig:trans}, along with the all the simulations. The in-situ observation points seem to be nicely replicated by the $ARCTCIGRIS$ run, whereas the $ARCTIC$ grid seems to biased positive in the higher-elevations of the ablation zone. The $f09$ grid is surprisingly competitive with the variable-resolution grids, capturing a realistic slope of the elevation-SMB curve. The elevation-SMB slopes of the uniform spectral-element and $f19$ grid simulate too shallow a slope in the higher elevation regions of the k-transect. Figure~\ref{ztrans} shows the representation of the surface of the ice sheet along these transects in the simulations, compared to a high resolution elevation dataset. The $1^{\circ}-2^{\circ}$ grids are noticeably coarse, with only a handful of grid cells populating the transect. The $f09$ grid is a bit of an exception, since the grid cells become very narrow in the meridional direction at high latitudes, and so a larger number of grid cells can populate the transect, consistent with its skillful representation of the ablation zone.

@@

\begin{figure}[t]
\begin{center}
         \includegraphics[width=130mm]{figs/temp_diffRACMO23p2_5p5km.pdf}
\end{center}
\caption{Climatological (1979-1998) annual precipitation rate bias (in mm/yr) relative to the RACMO2.3p2 5.5km resolution data product \cite{NETAL2019SCIENCE}.}
\label{fig:prect}
\end{figure}

>>>>

General Circulation Models (GCMs) are powerful tools for understanding the meteorology and climate of the high-latitudes, which are among the most sensitive regions on Earth to global and environmental change. Despite their importance, the numerical treatment of polar regions in GCMs is handled in vastly-different ways due to the so-called \textit{pole-problem} \cite{W2007JMSJ}, which refers to numerical instability arising from the convergence of meridians to polar singularities on latitude-longitude grids (e.g., Figure~\ref{fig:uni-grids}a). Depending on the numerics, methods exist to suppress this instability, and latitude-longitude grids may be advantageous for polar processes as structures can be represented with more degrees of freedom than elsewhere in the computational domain. With the recent trend towards globally uniform unstructured grids, any potential benefits of latitude-longitude grids on polar regions may become a relic of the past. In this study, a spectrum of model grids and dynamical cores (hereafter referred to as \textit{dycores}) available in the Community Earth System Model, version 2.2 (CESM; \url{http://www.cesm.ucar.edu/models/cesm2/}) are evaluated to understand their impacts on the simulated characteristics of the Arctic, with a special focus on the meteorology over the Greenland Ice Sheet.

@@

Initial simulations with the $ne30pg3$ spectral-element grid produced a weaker shortwave cloud forcing relative to the tuned up finite-volume dycore. All runs with the spectral-element dycore have two CLUBB parameter changes in order to provide a more realistic shortwave cloud forcing. These are CLUBB's $gamma$ parameter, reduced from 0.308 to 0.270, and c14, reduced from 2.2 to 1.6. Briefly, the $gamma$ parameter scales the width of the sub-grid distribution of vertical velocity, and c14 controls the strength of the damping term in the equation for the horizontal component of turbulent kinetic energy. For a thorough explanation of how CLUBB parameters impact the simulated climate, the reader is referred to \cite{GETAL2015JAMES}.

@@

All grids and dycores are run using an identical transient 1979-1998 AMIP-style configuration, with prescribed monthly SST/sea-ice after \cite{CESMSST}. This configuration refers to the $FHIST \ compset$ and runs out of the box in CESM2.2. The Community Land Model (CLM5; \url{https://escomp.github.io/ctsm-docs/versions/release-clm5.0/html/users_guide/index.html}), which uses the same grid as the atmosphere grid, calculates the surface energy balance at each land tile within a grid cell, which is used to compute snow and bare ice melting to inform the surface mass balance (SMB) of a glacier unit \cite{VKETAL2020JGR}. For land ice tiles coinciding with the Antarctic and Greenland Ice Sheets, the surface energy and mass balance are evaluated at ten elevation classes, and integrated over sub-grid hypsometric area-elevation bins for more accurate estimation of the surface mass balance of the big ice sheets. Ice accumulation is modeled as a capping flux, or snow in excess of the assumed 10 m snow cap, and refreezing of liquid within the snowpack additionally acts as a source of mass in the SMB calculation. Since the 10 m snowcap needs to be reached in the accumulation zone to simulate the SMB, the snow depths in the variable-resolution grids were spun-up by forcing CLM in standalone mode, cycling over 20 years of a fully coupled $ARCTIC$ run for about 500 years. The uniform resolution grids are all initialized with an SMB from an existing $f09$ spun-up initial condition.

@@

The SMB is the sum of the mass source term, accumulation (i.e., precipitation), and the mass sink term, ablation. Ablation can be expressed as evaporation/sublimation plus total runoff, with runoff being a combination of liquid precipitation and snow and ice melt. Not all liquid precipitation becomes runoff; rain may penetrate pore spaces in the firn layer and freeze, forming ice lenses in the subsurface. All these processes are represented by different components in CESM, and the surface mass balance is computed by CLM. CLM uses a downscaling technique to account for sub-grid variability in SMB. In short, the ice sheet patch in a CLM grid cell is subdivided into 10 elevation classes (EC), weighted by their respective area fractions at each EC, which is derived from a high resolution GrIS elevation model. The near surface air temperature, humidity and air density are calculated at each EC using an assumed lapse rate and the elevation difference from the grid mean, allowing for the precipitation rates to be repartitioned into solid or liquid based on the temperature of each EC. This technique allows for a unique surface energy balance and surface mass balance at each EC. Integrating over all ECs using the area weights then provides a more accurate SMB. For a detailed description of how the SMB is computed in the CESM, the reader is referred to \cite{SETAL2019TC,KETAL2020JAMES}.

@@

Simulated SMB are compared against several observational data products. RACMO2.3 11km and RACMO2.3p2 5.5km are regional model simulations that have been forced by reanalysis at its domain boundaries. The RACMO simulations have been shown to performs very skillfully against observations and is therefore considered an ideal modeling target \cite{NETAL2015TC,NETAL2019SCIENCE}. The Land Ice Verification and Validation toolkit (LIVVkit), version 2.1 \cite{LIVVkit} maintains a repository of snow pit and and ice core SMB measurements, as well as the IceBridge radar accumulation dataset. The LIVVkit dataset is compared against model simulations by finding the nearest grid cell center to the location of each observation.

A common high resolution dataset is used to generate the GrIS boundary conditions using the CLM dataset creation tools. Since we are interested in the total ice sheet SMB, we seek to integrate various components of the SMB over a common ice mask to get the total mass change of the GrIS. Figure~\ref{fig:grisdx} shows the GrIS ice mask area across the different grids, as a function of the number of grid points. Due to the use conservative regridding in the CLM tools, the interpolation errors are small and ice mask areas have less than $1.5\%$ errors relative to the raw ice mask dataset. RACMO2.3, however, uses a smaller ice mask, about $3\%$ smaller than the raw ice mask dataset. While Figure~\ref{fig:grisdx} suggests integrating quantities over the native ice mask of the six grids would probably not suffer from large errors due to differing ice masks, we seek to compare these integrated quantities with RACMO. Therefore, we have taken the approach of mapping all model fields to the lowest resolution grids and integrating over the respective low resolution ice masks. Due to the sensitivity of mapping errors to grid coordinates (i.e., unstructured or structured), all quantities are evaluated on both the $f19$ and $ne30pg2$, the lowest resolution grid for both dycores considered in this study. In addition, two remapping algorithms are used; ESMF conservative and TempestRemap high-order, monotone algorithm. In all, each integrated quantity is evaluated (at most) four times to provide an estimate of uncertainty due to differences in grid coordinates and remapping algorithm.

@@

Before delving into the simulated characteristics of the Arctic, the global mean differences between the various grids and dycores are assessed. Figure~\ref{fig:dT-lores} shows plots of the 1979-1998 annual mean, zonal mean height expressed as differences between the uniform resolution grids and dycores. The $f09$ grid is warmer than the $f19$ grid, primarily in the mid-to-high latitudes and throughout the depth of the troposphere. This is a common response to increasing horizontal resolution in GCMs \cite{PS2002CD,RETAL2006JC}, and \cite{HK2020QJRMS} has shown that this occurs in CAM due to greater resolved vertical velocities that in turn, facilitate greater condensational heating in the macrophyiscs routine in CLUBB. The right columns in Figure~\ref{fig:dT-lores} supports this interpretation, which shows an increase in the climatological CLUBB heating in the mid-latitudes in the $f09$ grid.

@@

It's useful to understand summer temperature biases, instead of annual means, due to its control on ice/snow melt \cite{O2001JAM,HT2008Paleo}. Figure~\ref{fig:dThyps} shows the 1979-1998 lower troposphere summer temperature bias relative to ERA5. It is computed from the 500 hPa-1000 hPa geopotential thickness, solving for the layer mean virtual temperature using the hypsometric equation. The results generally track with the analysis of the zonal mean height plots; increasing resolution from $f19$ to $f09$ leads to a warmer climate, and the $1^{\circ}$ spectral-elements grids are warmer than the finite-volume grids. The summer temperatures in the finite-volume grids are persistently colder than ERA5 at high latitudes, whereas the $1^{\circ}$ spectral-element grids are warmer than ERA5 at only very high-latitudes, north of $80^{\circ}$. The $ARCTIC$ grid has similar summer temperatures to the $1^{\circ}$ spectral-element grids, except for a region extending from eastern N.Atlantic to northern Eurasia. The $ARCTICGRIS$ grid is warmer than the $ARCTIC$ grid over most of the Arctic region, but in particular west of Greenland in the Baffin Island/Hudson Bay region, extending westward to the northern coast of Alaska.

Some of these temperature anomalies may be related to summer shortwave cloud forcing differences across the different grids and dycores. Figure~\ref{fig:SWCF} shows the summer short-wave cloud forcing climatology in the runs. The uniform $1^{\circ}-2^{\circ}$ grids have similar cloud forcing, in the range of -85 to -90 W/m2 averaged north of $40^{\circ}$ latitude. The $ARCTIC$ grid has a weaker cloud forcing of about -80 W/m2, with less negative cloud forcing over the north pole, and the poleward side of the northern hemisphere land masses. The $ARCTCGRIS$ grid has much weaker cloud forcing at -70 W/m2, being systematically lower over most of the region plotted. While the cloud forcing to the west of Greenland is reduced in $ARCTCIGRIS$, there are also anomalous southerly winds in this region during summer (not shown), and so the large warm bias to the west of Greenland in Figure~\ref{fig:dThyps} is likely a combination of both cloud radiative and circulation changes.

\begin{figure}[t]
\begin{center}
         \includegraphics[width=100mm]{figs/temp_contours_diffERA5_Thyps.pdf}
\end{center}
\caption{.}
\label{fig:dThyps}
\end{figure}

\begin{figure}[t]
\begin{center}
         \includegraphics[width=100mm]{figs/temp_contours_diffCERES_SWCF.pdf}
\end{center}
\caption{.}
\label{fig:SWCF}
\end{figure}

\begin{figure}[t]
\begin{center}
         \includegraphics[width=130mm]{figs/temp_contours_diffCERESdiffRACMO_FSDS.pdf}
\end{center}
\caption{.}
\label{fig:FSDS}
\end{figure}

@@

Figure~\ref{fig:pointdiff} shows the distribution of point-wise differences from LIVVkit observational database, as a violin plot. The IceBridge dataset is exclusively from the interior of the ice sheet and represents accumulation rates. The uniform $1^{\circ}-2^{\circ}$ grids have similar median errors of about +35-50 mm.w.e, while the variable-resolution errors are noticeably smaller. The in-situ observations in the accumulation zone, shown in the middle plot looks very similar to the IceBridge errors, and so provides confidence that the variable-resolution grids are outperforming the uniform grids in the interior accumulation zone. The errors evaluated at in-situ ablation zone measurements are in tension with the RACMO results in Figure~\ref{fig:tseries}. They indicate that the uniform $1^{\circ}-2^{\circ}$ grids perform similarly, the $ARCTIC$ grid showing a reduction in bias and the $ARCTICGRIS$ having the smallest bias of all grids. However, the in-situ ablation measurements are sparse in time and space, and so there is a large amount of uncertainty in extending these results to the entirety of the ablation zone.

The infamous k-transect mass balance profile is contained in the LIVVkit database, and shown in Figure~\ref{fig:trans}, along with all the simulations. The in-situ observation points seem to be nicely replicated by the $ARCTCIGRIS$ run, whereas the $ARCTIC$ grid seems to biased positive in the higher-elevations of the ablation zone. The $f09$ grid is surprisingly competitive with the variable-resolution grids, capturing a realistic slope of the elevation-SMB curve. The elevation-SMB slopes of the uniform spectral-element and $f19$ grid are too shallow, in particular at the higher elevation regions of the k-transect. Figure~\ref{ztrans} shows the representation of the surface of the ice sheet along these transects in the simulations, compared to a high resolution elevation dataset. The $1^{\circ}-2^{\circ}$ grids are noticeably coarse, with only a handful of grid cells populating the transect. The $f09$ grid is a bit of an exception. The $f09$ grid cells become very narrow in the meridional direction at high latitudes, and so a larger number of grid cells can populate the east-west transect, consistent with its skillful representation of the ablation zone compared with RACMO.

@@

\begin{figure}[t]
\begin{center}
         \includegraphics[width=130mm]{figs/temp_contours_diffCERESdiffRACMO_CLOUD_PRECIP.pdf}
\end{center}
\caption{Climatological (1979-1998) annual precipitation rate bias (in mm/yr) relative to the RACMO2.3p2 5.5km resolution data product \cite{NETAL2019SCIENCE}.}
\label{fig:prect}
\end{figure}
